---
title: "615-midterm"
author: "Chenghao Xia"
format: html
engine: knitr
---

```{r}
#| label: Load libraries
#| warning: false
#| message: false
#| echo: false
library(dplyr)
library(tidyverse)
library(esquisse)
library(rfema)
library(knitr)
library(kableExtra)
library(patchwork)
library(gridExtra)
```

```{r}
#| echo: false
d_2020<-read.csv("details_2020.csv",header = TRUE)
d_2021<-read.csv("details_2021.csv",header = TRUE)
fema<-read.csv("DisasterDeclarationsSummaries.csv",header = TRUE)
```
## Data Cleaning 
### Detail Document
We start data cleaning with Detail table.

First, we identify the type of event that relates to the flood. We chose "Flash Flood", "Flood", "Coastal Flood" and "Lakeshore Flood" as our new event_type variable.
```{r}
#| echo: false
event_types <- c(unique(d_2020$EVENT_TYPE))
related <- event_types[grepl("Flood", event_types)]
related_var <- c(related)
tidy_d2020<-d_2020 %>%
  filter(EVENT_TYPE %in% related_var)
```
Then we check if all the data here has the year equal to 2020/2021. And we prove that all the values are in 2020/2021.
```{r}
#| echo: false
not_2020<-any(d_2020$YEAR !=2020)
```

```{r}
#| echo: false
b_na <- colSums(is.na(tidy_d2020) | tidy_d2020 == "") == nrow(tidy_d2020)
col_na <- names(b_na[b_na])
tidy1_d2020 <- tidy_d2020[, !(names(tidy_d2020) %in% col_na)]
```
Finally, we find "Damage_Property" and "Damage_crops" should be changed into numeric for future calculation, then we remove "K" and times 1000 for the value.
```{r}
#| echo: false
convert_to_numeric <- function(value) {
  if (grepl("K", value)) {
    numeric_value <- as.numeric(sub("K", "", value))
    numeric_value * 1000
  } else if (grepl("M", value)) {
    numeric_value <- as.numeric(sub("M", "", value))
    numeric_value * 1000000
  } else {
    as.numeric(value)
  }
}
tidy1_d2020$DAMAGE_PROPERTY <- sapply(tidy1_d2020$DAMAGE_PROPERTY, convert_to_numeric)
tidy1_d2020$DAMAGE_CROPS <- sapply(tidy1_d2020$DAMAGE_CROPS, convert_to_numeric)
```

```{r}
#| echo: false
#Doing the same Data cleaning with Year 2021.
event1_types <- c(unique(d_2021$EVENT_TYPE))
related1 <- event1_types[grepl("Flood", event1_types)]
related1_var <- c(related1)
tidy_d2021<-d_2021 %>%
  filter(EVENT_TYPE %in% related1_var)
not_2021<-any(tidy_d2021$YEAR !=2021)
b1_na <- colSums(is.na(tidy_d2021) | tidy_d2021 == "") == nrow(tidy_d2021)
col1_na <- names(b1_na[b1_na])
tidy1_d2021 <- tidy_d2021[, !(names(tidy_d2021) %in% col1_na)]
tidy1_d2021$DAMAGE_PROPERTY <- sapply(tidy1_d2021$DAMAGE_PROPERTY, convert_to_numeric)
tidy1_d2021$DAMAGE_CROPS <- sapply(tidy1_d2021$DAMAGE_CROPS, convert_to_numeric)
```
### FEMA document
And now we start data cleaning with Fema Documents.

First we choose "Flood" in "incidentType" 

And then find the year which falls into 2020 and 2021. 
```{r}
#| echo: false
tidy_fema<-fema %>%
  filter(incidentType %in% "Flood") %>%
  filter(fyDeclared%in% c(2020,2021))
```
## EDA

### Flood distribution
First, we have a bar plot to show the appearance of these four event type in order to see the distribution  
```{r}
#| echo: false
E_T <- data.frame(
  VariableName = c("Flash Flood", "Flood", "Coastal Flood", "Lakeshore Flood"),
  Description = c("A sudden and rapid flooding of an area","When water covers normally dry land","When water from the ocean or sea rises and inundates coastal areas","A flooding event that specifically affects areas near the shores of lakes")
)
E_T_table <- kable(E_T, "simple")
print(E_T_table)
```

```{r}
#| echo: false
distr_2020<-ggplot(tidy1_d2020) +
  aes(x = EVENT_TYPE) +
  geom_bar(fill = "skyblue") +
  labs(
    x = "Type of Flood",
    y = "Counts",
    title = "Event Type Distribution for 2020"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
distr_2021<-ggplot(tidy1_d2021) +
  aes(x = EVENT_TYPE) +
  geom_bar(fill = "skyblue") +
  labs(
    x = "Type of Flood",
    y = "Counts",
    title = "Event Type Distribution for 2021"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14L, face = "bold",hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
grid.arrange(distr_2020,distr_2021, ncol = 2)
```
It is clear that "Flash Flood" has a dominant position, and "Flood" also has a much higher value compared to "Coastal Flood" and "Lakeshore Flood". But it is also clear that the total count of floods in 2021 is much bigger than the total count of floods in 2020.

### Injuries and Deaths
Then we consider which flood will bring the most injuries and deaths.

First, we may introduce injuries and deaths.
```{r}
#| echo: false
I_D <- data.frame(
  VariableName = c("Injuries_Direct(ID)","Injuries_indirect(II)","Deaths_Direct(DD)","Deaths_Indirect(DI)"),
  Description = c("The number of people get hurt dircetly because of weather event","The number of people get hurt indirectly because of weather event","The number of people die directly because of weather event","The number of people die indirectly because of weather event")
)
I_D_table <- kable(I_D, "simple")
print(I_D_table)
```

```{r}
#| echo: false
ID_2020<-ggplot(tidy1_d2020) +
  aes(x = EVENT_TYPE, weight = INJURIES_DIRECT) +
  geom_bar(fill = "skyblue") +
  labs(
    x = "Type of Flood",
    y = "Count for ID",
    title = "ID number for 2020"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
II_2020<- ggplot(tidy1_d2020) +
  aes(x = EVENT_TYPE, weight = INJURIES_INDIRECT) +
  geom_bar(fill = "skyblue") +
  labs(
    x = "Type of Flood",
    y = "Count for II",
    title = "II number for 2020"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
DD_2020<-ggplot(tidy1_d2020) +
  aes(x = EVENT_TYPE, weight = DEATHS_DIRECT) +
  geom_bar(fill = "skyblue") +
  labs(
    x = "Type of Flood",
    y = "Count for DD",
    title = "DD number for 2020"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
DI_2020<-ggplot(tidy1_d2020) +
  aes(x = EVENT_TYPE, weight = DEATHS_INDIRECT) +
  geom_bar(fill = "skyblue") +
  labs(
    x = "Type of Flood",
    y = "Count for DI",
    title = "DI number for 2020"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

```{r}
#| echo: false
ID_2021<-ggplot(tidy1_d2021) +
  aes(x = EVENT_TYPE, weight = INJURIES_DIRECT) +
  geom_bar(fill = "skyblue") +
  labs(
    x = "Type of Flood",
    y = "Count for ID",
    title = "ID number for 2021"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
II_2021<- ggplot(tidy1_d2021) +
  aes(x = EVENT_TYPE, weight = INJURIES_INDIRECT) +
  geom_bar(fill = "skyblue") +
  labs(
    x = "Type of Flood",
    y = "Count for II",
    title = "II number for 2021"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
DD_2021<-ggplot(tidy1_d2021) +
  aes(x = EVENT_TYPE, weight = DEATHS_DIRECT) +
  geom_bar(fill = "skyblue") +
  labs(
    x = "Type of Flood",
    y = "Count for DD",
    title = "DD number for 2021"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
DI_2021<-ggplot(tidy1_d2021) +
  aes(x = EVENT_TYPE, weight = DEATHS_INDIRECT) +
  geom_bar(fill = "skyblue") +
  labs(
    x = "Type of Flood",
    y = "Count for DI",
    title = "DI number for 2021"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

```{r}
#| echo: false
grid.arrange(ID_2020,ID_2021, ncol = 2)
```
For Injuries_Direct, "Flash Flood" has a dominant position over others in both 2020 and 2021.
```{r}
#| echo: false
grid.arrange(II_2020,II_2021, ncol = 2)
```
Since the amount of Injuries_Indirect is not a lot, we can not say there is a big difference between 2020 and 2021 though the distribution is different.
```{r}
#| echo: false
grid.arrange(DD_2020,DD_2021, ncol = 2)
```
For Deaths_Direct, "Flash Flood" has a dominant position over others in both 2020 and 2021.
```{r}
#| echo: false
grid.arrange(DI_2020,DI_2021, ncol = 2)
```
Since the number of indirect injuries is not large, we cannot say there is a big difference between 2020 and 2021, though the distribution is different.

As a short summary, the distributions of injuries and deaths for 2020 and 2021 are pretty similar. For direct injuries and deaths, we can make the hypothesis that "flash flood" is most common but dangerous since it has a dominant position among all four types of flood. For indirect injuries and deaths, we cannot draw any conclusions since the total count is too small and the distribution is different in 2020 and 2021.

### Flood Cause
Then we will consider how floods are distributed in order to prevent them
```{r}
#| echo: false
FC_2020 <- table(tidy1_d2020$FLOOD_CAUSE)
FC_2021 <- table(tidy1_d2021$FLOOD_CAUSE)
perc_2020 <- (FC_2020 / sum(FC_2020)) * 100
perc_2021 <- (FC_2021 / sum(FC_2021)) * 100
data.frame(P_2020= perc_2020)
```

```{r}
#| echo: false
data.frame(P_2021= perc_2021)
```
Compared to 2020 and 2021, "heavy rain" has a dominant position in the flood cause. It does make sense since heavy rain is the most common event that will happen in our daily lives. But it is hard for the government to prevent floods. If the government finds out there is heavy rain, they will not conclude there will be a flood.

### Damage to property or crops
Since the flood cause cannot help us prevent the flood, we are going to find the relationship between month and damage property or damage crop.

If there is a relationship between month and damage cost, the government can focus more attention on the high-cost month.
```{r}
#| echo: false
Damage <- data.frame(
  VariableName = c("Damage of Property","Damage of Crops"),
  Description = c("The estimated amount of damage to property incurred by the weather event","The estimated amount of damage to crops incurred by the weather event")
)
Damage_table <- kable(Damage, "simple")
print(Damage_table)
```

```{r}
#| echo: false
months<-c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")
tidy1_d2020$MONTH_NAME<-factor(tidy1_d2020$MONTH_NAME,levels=months,ordered=TRUE)
tidy1_d2020<-tidy1_d2020[order(tidy1_d2020$MONTH_NAME), ]
tidy1_d2021$MONTH_NAME<-factor(tidy1_d2021$MONTH_NAME,levels=months,ordered=TRUE)
tidy1_d2021<-tidy1_d2021[order(tidy1_d2021$MONTH_NAME), ]
pro_2020<-ggplot(tidy1_d2020) +
  aes(x = MONTH_NAME, weight = DAMAGE_PROPERTY) +
  geom_bar(fill = "black") +
labs(
    x = "Number of Month",
    y = "Damage of Property",
    title = "Damage of Property in 2020"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
pro_2021<-ggplot(tidy1_d2021) +
  aes(x = MONTH_NAME, weight = DAMAGE_PROPERTY) +
  geom_bar(fill = "black") +
labs(
    x = "Number of Month",
    y = "Damage of Property",
    title = "Damage of Property in 2021"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
crop_2020<-ggplot(tidy1_d2020) +
  aes(x = MONTH_NAME, weight = DAMAGE_CROPS) +
  geom_bar(fill = "black") +
labs(
    x = "Number of Month",
    y = "Damage of Crops",
    title = "Damage of Crops in 2020"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
crop_2021<-ggplot(tidy1_d2021) +
  aes(x = MONTH_NAME, weight = DAMAGE_CROPS) +
  geom_bar(fill = "black") +
labs(
    x = "Number of Month",
    y = "Damage of Crops",
    title = "Damage of Crops in 2021"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

```{r}
#| echo: false
grid.arrange(pro_2020,pro_2021, ncol = 2)
```

We can see the damage to property has a few similarities between 2020 and 2021. They both have a high value for September. But in May 2020, it had a high flood value, which didn't happen in May 2021. Similar to that, in August 2021, it has a low value, which August 2020 doesn't have.
```{r}
#| echo: false
grid.arrange(crop_2020,crop_2021, ncol = 2)
```
We can see the damage to crops is similar for both 2020 and 2021, since July and August are two near months. We can make the hypothesis that in July and August, the damage to crops is the most severe among the whole year.

### State
By looking at the cost per month, we want to find which states experience the lot floods.

First, we have histograms with the top ten values of flood states
```{r}
#| echo: false
count_2020=table(tidy1_d2020$STATE)
most_2020=head(sort(count_2020,decreasing=TRUE),10)
count_2021=table(tidy1_d2021$STATE)
most_2021=head(sort(count_2021,decreasing=TRUE),10)
data<-data.frame(List1=most_2020,List2=most_2021)
```

```{r}
#| echo: false
data_2020<-ggplot(data) +
  aes(x=List1.Var1,y = List1.Freq) +
  geom_col(fill = "red") +
labs(
    x = "State's name",
    y = "Counts",
    title = "States with flood in 2020"
    ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
    caption = "Choose top ten States"
data_2021<-ggplot(data) +
  aes(x=List2.Var1,y = List2.Freq) +
  geom_col(fill = "red") +
labs(
    x = "State's name",
    y = "Counts",
    title = "States with flood in 2021",
    caption = "Choose top ten States"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
grid.arrange(data_2020,data_2021, ncol = 2)
```
It is interesting since the number of states with floods is totally different between 2020 and 2021. Maybe we can make the hypothesis that some states often get a lot of floods, but for most states, the floods are random, and it is hard to predict whether the next year will be more or less.

## FEMA EDA
First we make some explanation of these variables
```{r}
#| echo: false
P_D <- data.frame(
  VariableName = c("ihProgramDeclared", "iaProgramDeclared", "paProgramDeclared	", "hmProgramDeclared"),
  Description = c("Whether the Individuals and Households program was declared for this disaster","Whether the Individual Assistance program was declared for this disaster","Whether the Public Assistance program was declared for this disaster","Whether the Hazard Mitigation program was declared for this disaster")
)
P_D_table <- kable(P_D, "simple")
print(P_D_table)
```

```{r}
#| echo: false
#| message: false
tidy1_fema<- tidy_fema %>%
  select(ihProgramDeclared,iaProgramDeclared,paProgramDeclared,hmProgramDeclared)
tidy2_fema <- reshape2::melt(tidy1_fema)
ggplot(tidy2_fema, aes(x = variable, fill = factor(value))) +
  geom_bar(position = "dodge") +
  labs(
    title = "Declaration of Program", 
    x = "Variable", 
    y = "Counts") +
  scale_fill_manual(values = c("1" = "blue", "0" = "red"))+
  theme(
    plot.title = element_text(size = 12L, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 11L, face = "bold"),
    axis.title.x = element_text(size = 11L, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```
We can see Individual Assistance program didn't declared in the flood during 2020 and 2021, while Public Assistance program was declared a little time. Hazard Mitigation program was declared almost every time

## Summary
For the flood during 2020 and 2021, "flash flood" and "flood" are the main parts of the event. But "Flash Flood" is the most dangerous since more than 80% of people were hurt or died because of it. And the main cause of floods is "heavy rain," which accounts for approximately 90% of all the causes. It does make sense since "heavy rain" is the most common event we can meet every day.

Looking at the damage to property and crops, August and September are two months in which property and crops will be damaged. Although there are some abnormal values between 2020 and 2021, the overall distribution still has a lot of similarities. But when we compare the top 10 states that experienced the most floods, there is a big difference between 2020 and 2021. So we can make the hypothesis that the flood will occur in different states over the years.

With the Fema data, we can see the Individual Assistance Program wasn't declared in the flood during 2020 and 2021, which means this program has a lot of work to do compared to the Hazard Mitigation Program, which was declared almost every time. The other two programs had some differences but did work in some case.